$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in the Presence of Non-Functional Properties

### 2014-03-06 JournalToSubmit2013/modelingWithPISODM.tex
New organization of the paper, new version of section 3 and new section 4. Introduction and related works reviewed 06.04.2014

This section presents $\pi$SOD-M, an MDD-based methodology for building service compositions with non-functional requirements. 
$\pi$SOD-M extends SOD-M with concepts for modelling non-functional requirements, it proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models (Figure~\ref{fig:piSOD-M}).
\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
\caption{$\pi$SOD-M Overview.}
\label{fig:piSOD-M}
\end{figure}

@@

The e$^3$value model identifies the value/information exchange between components of the system. 
It defines \textit{dependency paths}, showing the value exchange between providers and end users once they ask for a service.
% , which are triggered by the occurrence of an end-consumer need.
A dependency path has a direction and consists of a sequence of linked dependency nodes.
A dependency path starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node\footnote{see Figure~\ref{fig:CIM:tpme3v}}. 
Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).

The BPMN model is a graphical representation that establishes the business process of the application through a high-level workflow. The next example illustrates the use of these models.

\begin{example}[To Publish Music]\label{ex:toPublicMusic}
To illustrate our proposal, consider the scenario ``To Publish Music''  based on
three external actors ({\em Spotify, Twitter} and {\em Facebook}), as well as the user and the application itself. 
The e$^3$value and BPMN models for this application are shown in Figures~\ref{fig:CIM:tpme3v} and \ref{fig:CIM:tpmbpmn}. 
\end{example}

\begin{figure}[htpb]
\center
\includegraphics[width=0.88\textwidth]{figs/e3value.pdf}
\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
\end{figure}

The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed. 
The E3 value model in Figure \ref{fig:CIM:tpme3v} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users. The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription). Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).

%Note that the models at this level show\dots 
%{\color{red} Valeria and Placido: Please explain the CIM models for this example.}
%\hfill\openbox

Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
%
\begin{figure}[htpb]
\center
\includegraphics[width=0.49\textwidth]{figs/SC.pdf}
\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
\end{figure}

The CIM level models are the basis for the development of the application. 
The information presented at this level are (manually) refined into PIM-level models.

@@

The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements. 
Non-functional requirements are defined as \textit{constraints} over actions and data. The \textit{$\pi$-UseCase} meta-model describes the functional aspects of the application being developed and the restrictions that applies to each of them. These restrictions are modeled using the concepts: (i) constraint type and (ii) non-functional requirement with its specific attribute.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
%Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.

 We use part of our case study to illustrate this meta-model. The "To Publish Music" use case can update the Facebook or Twitter user status. Therefore, it is necessary to perform a social network authentication with the users data. Each social network uses different services and different forms of authentication. The authentication constraint is required to update a music status. The restriction is stereotyped as a «value» constraint, because the users id and password are verified.  Figure \ref{fig:CIM:tpmbpmn}  the buy music, download music, listen music and pay use cases. The process of buying a song requires the user private data for a Spotify account, and also a secure connection, represented as «value» and «business» constraint, respectively. For payment, the user must provide the data of payment card or PayPal account login and password, represented as «value» stereotype . Other restriction is that the minimum payment value is 2 euros.
%{\color{red} Talk about this model for the example.}
\end{example}

@@

 All constraints are described by stereotyped use cases. Every relationship between a use
case and a constraint is done through a dependency relationship (from UML).  A constraint is defined by a stereotype, containing its type and name. Constraints are represented by a stereotyped use case («constraint»).
Each use case is related with one or more system requirements. Moreover, these use cases are related to non-functional requirements necessary for the proper execution of the application function. The function described by a use case can be part of a business service (in the case of service remote calls). This information is described through comments.

The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent constraints over data and actions 
gathers the constraints described in the \textit{$\pi$-UseCase}. A \textit{$\pi$-ServiceProcess} model represents the service processes and its contracts. It models the set of system activities and describe contracts for each activity or the process as a whole. Each service identified in the previous model (\textit{$\pi$-UseCase}) is mapped to an action. This model describes the workflow for the modelled application behaviour. The behaviour is represented by each action and the restrictions by the assertion description. All assertions are described by stereotyped actions.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
Considering the example scenario, the contract based process of activities execution is shown in figure \ref{fig:CIM:serviceprocess}. The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service. The buy music pre-conditions are: (i) verify if the User data are correct; (ii) if the User is already logged in Spotify; (iii) if bank account information are correct and; (iv) if there is enough money to make the payment. As post-condition, it ensures the complete transaction and verify if a notification were sent to the user and Spotify, about the payment authorization. There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified. To update services, depending of each service, there may be different restrictions. As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in case of Twitter. In the case of Facebook, it is required that the user is already logged on Spotify and these data are the same as Facebook. As post-condition, it ensures that the Facebook service send a notification of success. To update Twitter a pre-condition is required, while to update Facebook is necessary a pre-condition and a confirmation notice is modelled as post-condition. As a pre-condition for "twitter update" it is necessary that (i) the music format be correct and (ii) the twitter login and password be correct for the update.
%{\color{red} Talk about this model for the example.}
\end{example}

@@

The \textit{$\pi$-ServiceComposition} meta-model provides the concept  of \textit{Policy}
to integrate contracts with similar non-functional requirements.
For instance, security and privacy restrictions may be grouped into a security policy. As shown in Figure \ref{fig:e-scomposition-metamodel},
provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
Instances of this meta-model are represented as UML activity diagrams. 
In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
 white boxes represent classes modelling functional ones.

The {\em A-policy} based service composition ($\pi$-SCM) meta-model, 
%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
An instance of this meta-class is graphically represented as a partition in the activity diagram. 
A collaborator can be either internal or external to the system. 
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

@@

\item In order to represent constraints types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules. 
Intuitively, the conditions of each rule will be checked.
In case of no compliance, the actions defined by the rule will be performed.
The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
\end{itemizedTrivlist}

\begin{figure}[t]%[htpb]
\centering
\includegraphics[width=0.95\textwidth]{figs/e-composition-model}

{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}

\caption{$\pi$-SCM for the ``To publish music'' business service.}
\label{fig:servicecompositionmodel}
\end{figure}

>>>>

This section presents $\pi$SOD-M, an MDD-based methodology for building service compositions with non-functional requirements. 
$\pi$SOD-M provides concepts for modelling non-functional requirements, it proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models (Figure~\ref{fig:piSOD-M}).
\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
\caption{$\pi$SOD-M Overview.}
\label{fig:piSOD-M}
\end{figure}

@@

% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .   
\subsubsection{E$^3$value meta-model}
% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   . 
The e$^3$value model identifies the value/information exchange between components of the system. 
It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
% , which are triggered by the occurrence of an end-consumer need.
A dependency path has a direction and consists of a sequence of linked dependency nodes.
It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Figure~\ref{fig:CIM:tpme3v}). 
Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
The e$^3$value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed. 
\begin{figure}[htpb]
\center
\includegraphics[width=0.88\textwidth]{figs/e3value.pdf}
\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
\end{figure}

\begin{example}[To Publish Music]\label{ex:toPublicMusic}
To illustrate our proposal, consider the scenario ``To Publish Music''  based on
three external actors ({\em Spotify, Twitter} and {\em Facebook}), as well as the user and the application itself. 
The e$^3$value  model for this application are shown in Figures~\ref{fig:CIM:tpme3v}. The e$^3$value value model in Figure \ref{fig:CIM:tpme3v} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users. The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription). Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
\end{example}

%Note that the models at this level show\dots 
%{\color{red} Valeria and Placido: Please explain the CIM models for this example.}
%\hfill\openbox
% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   . 
\subsubsection{BPMN meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   . 
The BPMN meta-model is a graphical representation that establishes the business process of the application through a high-level workflow. The next example illustrates the use of this meta-model.

\begin{example}[To Publish Music]\label{ex:toPublicMusicBPMN}
Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
\end{example}
%
\begin{figure}[htpb]
\center
\includegraphics[width=0.49\textwidth]{figs/SC.pdf}
\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
\end{figure}

The CIM level models are the basis for the development of the application. 
The information presented at this level is (manually) refined into PIM-level models of the methodology $\pi$-SODM described next.

@@

 % .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   . 
\subsubsection{\textit{$\pi$-UseCase} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   . 
The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements. 
Non-functional requirements are defined as \textit{constraints} over actions and data. The \textit{$\pi$-UseCase} meta-model describes the functional aspects of the application being developed and the restrictions that apply to each of them. These restrictions are modeled using the concepts: (i) constraint type and (ii) non-functional requirement with its specific attribute. All constraints are described by stereotyped use cases. Every relationship between a use
case and a constraint is done through a dependency relationship (from UML).  A constraint is defined by a stereotype, containing its type and name. Constraints are represented by a stereotyped use case («constraint»).
Each use case is related with one or more system requirements. Moreover, these use cases are related to non-functional requirements necessary for the proper execution of the application function. The function described by a use case can be part of a business service (in the case of service remote calls). This information is described through comments.
We use part of our case study to illustrate this meta-model. The "To Publish Music" use case can update the Facebook or Twitter user status.
\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
%Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.

  Therefore, it is necessary to perform a social network authentication with the users data. Each social network uses different services and different forms of authentication. The authentication constraint is required to update a music status. The restriction is stereotyped as a «value» constraint, because the users id and password are verified.  Figure \ref{fig:CIM:tpmbpmn} shows the buy music, download music, listen music and pay use cases. The process of buying a song requires the user private data for a Spotify account, and also a secure connection, represented as «value» and «business» constraint, respectively. For payment, the user must provide the data of payment card or PayPal account login and password, represented as «value» stereotype. Other restriction is that the minimum payment value is 2 euros.
%{\color{red} Talk about this model for the example.}
\end{example}

@@

 % .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   . 
\subsubsection{\textit{$\pi$-ServiceProcess} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent constraints over data and actions 
groups the constraints described in the \textit{$\pi$-UseCase}. A \textit{$\pi$-ServiceProcess} model represents the service processes and its contracts. It models the set of system activities and describe contracts for each activity or the process as a whole. Each service identified in the previous model (\textit{$\pi$-UseCase}) is mapped to an action. This model describes the workflow for the modelled application behaviour. The behaviour is represented by each action and the restrictions by the assertion description. All assertions are described by stereotyped actions.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}. The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service. The buy music pre-conditions are: (i) verify if the User data are correct; (ii) if the User is already logged in Spotify; (iii) if bank account information are correct and; (iv) if there is enough money to make the payment. A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization. There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified. To update services, depending of each service, there may be different restrictions. As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in case of Twitter. In the case of Facebook, it is required that the user is already logged on Spotify and these data are the same as Facebook. As post-condition, it ensures that the Facebook service send a notification of success. To update Twitter a pre-condition is required, while to update Facebook is necessary a pre-condition and a confirmation notice is modelled as post-condition. As a pre-condition for "twitter update" it is necessary that (i) the music format be correct and (ii) the twitter login and password be correct for the update.
%{\color{red} Talk about this model for the example.}
\end{example}

@@

% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   . 
\subsubsection{\textit{$\pi$-ServiceComposition} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

The \textit{$\pi$-ServiceComposition} meta-model provides the concept   \textit{Policy}
to group contracts with similar non-functional requirements.
For instance, security and privacy restrictions may be grouped into a security policy. As shown in Figure \ref{fig:e-scomposition-metamodel}, $\pi$-SODM 
provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
Instances of this meta-model are represented as UML activity diagrams. 
In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
 white boxes represent classes modelling functional ones.

The {\em A-policy} based service composition ($\pi$-SCM) meta-model consists of:
%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
An instance of this meta-class is graphically represented as a partition in the activity diagram. 
A collaborator can be either internal or external to the system. 
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

@@

\item In order to represent constraint types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules. 
Intuitively, the conditions of each rule will be evaluated.
In case of no compliance, the actions defined by the rule will be performed.
The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
\end{itemizedTrivlist}

\begin{figure}[t]%[htpb]
\centering
\includegraphics[width=0.95\textwidth]{figs/e-composition-model}
%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
\caption{$\pi$-SCM for the ``To publish music'' business service.}
\label{fig:servicecompositionmodel}
\end{figure}

### 2014-03-06 JournalToSubmit2013/relworks.tex
New organization of the paper, new version of section 3 and new section 4. Introduction and related works reviewed 06.04.2014

A non-functional property (NFP) specifies a criterion about the behavior of a system related to the conditions in which it is executed and to its performance. Non-functional properties are also referred to as constraints, quality attributesÄù, quality goalsÄù, Äúquality of service requirements and Äúnon-behavioral requirementsÄù \cite{StGr05}. In the case of service-based applications, non-functional requirements concern the application itself as well as its component services. 
The majority of the work on NFPs uses a product-oriented approach concerned with measuring in which extent a software system is fulfills the NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.

Existing  approaches address NFP from four perspectives: (i) NFP expressed as quality of service properties (QoS attributes) that can be associated to service compositions; (ii) service matching for associating certain non-functional properties by composing specific services;  (iii) programming NFP as protocols that are asserted to service composition code; and, (iv) methodologies that provide tools for building service compositions with associated NFP.

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Classifying Non-functional properties}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
We conducted a systematic review \cite{placidoPhDThesis2012} to analyze approaches that consider non-functional properties for developing service-oriented applications. As a result this analysis proposes a classification of NFP as a result of a study concerning software methodologies for the construction of service-oriented systems. The classification is organized in three layers: application modeling, services composition and services.

\begin{itemize}
\item At the application layer NFP  refer to business rules (e.g., only the user can publish data on her/his wall) and values for example, the email address is a string containing an "@"Äù and a "Äú."Äù. A value NFR expresses constraints about the way data and functions can be accessed and executed.

\item The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements (NFP).

\item Business NFP at the service layer concern properties that are associated to services and defines how to call their exported operations (business properties).  Value constraints concern more on the conditions in which services can be used. For example, accessing to a function within an authentication protocol. This layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.  For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer. 
\end{itemize}

@@

\cite{CyPr04} proposes an extension to UML for expressing NFPs that can be integrated into the Class, Sequence, and Collaboration Diagrams.    UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram are annotated with quantitative performance constraints. The UML Profile for Modeling Quality of Service and Fault Tolerance Characteristics and Mechanisms \cite{Omg04} describe a set of QoS characteristics such as throughput, latency, integrity, and dependability. The characteristics can also be parameterized with for example the type of keys or types of encryption algorithms in case of the security characteristics. For all UML profiles based approaches it is  possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g. Class or Operation. For instance, there are stereotypes for encryption, decryption and logging which can be added to classes or operations that are representing services. The ordering or priority of stereotypes cannot be expressed but composition-related requirements can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams \cite{OrHe07}.

@@

%Business rules provide means for expressing, managing and updating pieces of busi-ness domain knowledge independent of the rest of the application. According to the Business Rules Group \cite{Hay00}, a business rule  is intended to assert business structure or to control the behavior of the business \cite{Hay00}. Business rules are usually expressed either as constraints or in the form if conditions then action. The conditions are also called rule premises. The business rule approach encompasses a collection of terms (definitions), facts (connection between terms) and rules (computation, constraints and conditional logic) \cite{Hall01}. Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making. 
In a business rule system,  rules are logically separated   from the system \cite{ChMe04}. \cite{ChMe04} investigate a hybrid approach, that combines business processes as known in BPEL with business rules \cite{Hay00}.

\cite{ChLa09} proposed a generative framework to address NFPs for web service compositions. They separate the orchestration meta-model from other non-functional meta-models. They propose a non-functional meta-model for security  and they specify how these requirements must be ensured. The NFPs are  associated to the composition model by adding annotations.
KAOS \cite{DaLF93,Lams01} is a goal-oriented framework for addressing functional requirements with a formal temporal logic  to precisely specify NFPs, such as performance, as part of attribute specification of functional requirements.

@@

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Towards methodologies for developing Service-oriented applications   with non-func\-tion\-al properties}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -

>>>>

A non-functional property (NFP) specifies a criterion about the behavior of a system related to the conditions in which it is executed and to its performance. Non-functional properties are also referred to as constraints, quality attributesÄù, quality goalsÄù, Äúquality of service requirements and Äúnon-behavioral requirementsÄù \cite{StGr05}. In the case of service-based applications, non-functional requirements concern the application itself as well as its component services. 
The majority of the works on NFPs focus on measuring in which extent a software system  fulfills the NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.

We consider that existing  approaches address NFP from four perspectives: (i) NFP classification generally expressed as quality of service properties (QoS attributes) that can be associated to service compositions; (ii) service matching for satisfying  non-functional properties by composing specific services;  (iii) programming NFP as protocols that are asserted to service composition code; and, (iv) methodologies that provide tools for building service compositions with associated NFP.

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Classifying non-functional properties}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
We conducted a systematic review \cite{placidoPhDThesis2012} to analyze approaches that classify non-functional properties for developing service-oriented applications. As a result this analysis proposes a classification of NFP.
%as a result of a study concerning software methodologies for the construction of service-oriented systems. 
The classification is organized into three layers: application, services composition and services.

\begin{itemize}
\item At the application layer NFP  refer to business rules (e.g., only the user can publish data on her/his Facebook wall) and values for example, the email address is a string containing an "@"Äù and a "Äú."Äù. A value NFP expresses constraints about the way data and functions can be accessed and executed.

\item The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements.

\item  At the service layer  business NFP concern properties that are associated to services and defines how to call their exported operations (business properties).  Value constraints concern more on the conditions in which services can be used. For example, accessing to a function within an authentication protocol. This layer gives an abstract view of the kind of properties associated to services that can be combined for providing NFP for a composition.  For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer. 
\end{itemize}

@@

\cite{CyPr04} proposes an extension to UML for expressing NFPs that can be integrated into the Class, Sequence, and Collaboration Diagrams.    UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram are annotated with quantitative performance constraints. The UML Profile for Modeling Quality of Service and Fault Tolerance Characteristics and Mechanisms \cite{Omg04} describe a set of QoS characteristics such as throughput, latency, integrity, and dependability. The characteristics can also be parameterized with for example the type of keys or types of encryption algorithms to used for addressing  security. For all UML profiles based approaches it is  possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g. Class or Operation. For instance, there are stereotypes for encryption, decryption and logging which can be added to classes representing operations exported by services. The ordering or priority of stereotypes cannot be expressed but composition-related requirements can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams \cite{OrHe07}.

@@

%Business rules provide means for expressing, managing and updating pieces of busi-ness domain knowledge independent of the rest of the application. According to the Business Rules Group \cite{Hay00}, a business rule  is intended to assert business structure or to control the behavior of the business \cite{Hay00}. Business rules are usually expressed either as constraints or in the form if conditions then action. The conditions are also called rule premises. The business rule approach encompasses a collection of terms (definitions), facts (connection between terms) and rules (computation, constraints and conditional logic) \cite{Hall01}. Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making. 
%In a business rule system,  rules are logically separated   from the system \cite{ChMe04}. \cite{ChMe04} investigate a hybrid approach, that combines business processes as known in BPEL with business rules \cite{Hay00}.

\cite{ChLa09} proposes a generative framework to address NFPs for web service compositions. They separate the orchestration meta-model from other non-functional meta-models. They propose a non-functional meta-model for security  and they specify how these requirements must be ensured. The NFPs are  associated to the composition model by adding annotations.
KAOS \cite{DaLF93,Lams01} is a goal-oriented framework for addressing functional requirements with a formal temporal logic  to precisely specify NFPs, such as performance.
% as part of attribute specification of functional requirements.

@@

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{ Methodologies for developing Service-oriented applications   with non-func\-tion\-al properties}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
